A molecular taxonomy of tumors independent of tissue-of-origin

### 2020-08-16 text/discussion.Rmd
edit full document

To our knowledge, this approach has not previously been attempted in the
literature in spite of its relative simplicity. We limited our analysis to
mutations with likely deleterious effects, including missense, nonsense,
frame-shift, stop-gain, untranslated region and splice site mutations in genes that are actively
expressed in each cancer type, thus avoiding bias from transcription coupled
repair. Our method of filtration differs from the "rank-and-cut" method of
[@PCAWG-MAIN] but represents a reasonable attempt to account for the same
biases. We also filtered genes on the basis of pathway membership in biochemical
pathways, excluding curated gene sets related to diseases, syndromes, or classes
of proteins with shared catalytic activity or conserved domains ( _e.g._ transcription factors) which are potentially problematic
[@pathways-Atul-Butte]. We chose this approach to limit redundancy, which is
already high in the remaining pathways for some driver genes, and to exclude
biologically unrelated collections of genes. We do not account for non-coding
mutations, amplification events or structural variants, though we hope to
include them in subsequent studies (see below). Our pathway-centric analysis
also allows us to circumvent the problem that some cancer-promoting mutations
likely occur at low enough frequencies as to be indistinguishable from
background. Previous attempts to capture these associations using
Bayesian networks have yielded important new insights into this problem and new
genes [@hornNetSigNetworkbasedDiscovery2018] that would not have been possible
with traditional analysis, supporting the case that pathway-centric approaches
have the potential to increase statistical power.

@@

An alternative hypothesis is that every tumor belongs to one of a large number
of syndromes which are unique to each tissue-of-origin, that share some
mechanisms and treatment strategies. Recent publication of TCGA consortium
papers present a view largely, and surprisingly, consistent with this hypothesis
[@hoadleyMultiplatformAnalysis2014; @hoadleyCellofOriginPatternsDominate2018].
Perhaps owing to the intractable complexity of genomics, proteomics, and
patient metadata in all its forms, the inescapable conclusion thus far is that
tissue-of-origin remains the most important driver of tumor characteristics at
every scale and by every measure. Our observations contrast with this view, and
instead support an interpretation of publicly available data in which all tumors
manifest one of a limited number of metastable phenotypes resulting from
disruptions in basic pathways represented by our classes (**figure 1**).

@@

Our results illustrate how unique combinations of mutations in pan-cancer driver genes with tissue-specific pathway disruptions result in common categories when viewed at the level of the pathway knowledge-base. For example, we find that for "snRNP assembly", one of the unique driver pathways of cluster 5, the top gene in breast cancer is represented by TPR (mutated in 26% of samples). By contrast the top gene for head and neck and lower grade gliomas is RANBP (20% and 13% of samples, respectively). For some other clusters, _e.g._ 7, 8 & 9, the top genes are known drivers such as TP53, PIK3CA and KRAS. None of these genes is 100% mutated in any cluster however, and therefore it is the combinations with secondary mutations, many of which are tissue-specific, that round out the complement of pathways in all clusters. Some top genes are found in most of the clusters (supplemental figure S5), in spite of the fact that they contribute to many cluster-specific pathways. This can only be explained as a result of other less common driver genes complementing the pathway disruption of the unique combination of driver genes that are disrupted in each tumor, and we speculate that many of these are relatively tissue-specific in terms of their sensitivity to mutation. Consistent with our observations, @moonlight showed how tissue- and cancer-specific driver genes and druggable targets of pathway moduli can be discovered by machine learning on integrated multi-omics datasets. Marrying these approaches in future work will greatly enhance our understanding of tumor biology.

@@

By modeling CDC longevity data as a baseline risk function we showed that each
of our pathway disruption classes exhibit cancer-type specific effects on
expected survival (**figure 7**). Contrary to what one might expect at the
outset, we found that models assuming different survival based on class and
cancer type fit better than models in which they are independent
factors. However, considering that within each cancer type there are different
clinical standards of care, and even within classes of drugs the preferred
treatment can very between cancers, it makes sense that we observe tissue-specific class effects. Contrast the situation with ovarian _vs._ breast cancer, which
are both hormonally driven cancers, for example. Ovarian cancer has but one main
treatment axis, platinum, whereas breast cancer patients have a variety of
treatment regimens based on molecular subtype and other factors. It is easy to
see how standard of care for one cancer and class could produce different
outcomes than the same class of a different cancer type. Unfortunately, given
the diversity drug classes and treatments (including non-chemical treatments
such as surgery and palliative care), we lack sufficient power 
to explore these variables in the TCGA data. It is our hope that 
future studies will help to distinguish between treatment-specific
effects on survival given different pathway disruption classes.

>>>>

To our knowledge, this approach has not previously been attempted in the
literature in spite of its relative simplicity. We limited our analysis to
mutations with likely deleterious effects, including missense, nonsense,
frame-shift, stop-gain, untranslated region and splice site mutations in genes
that are actively expressed in each cancer type, thus avoiding bias from
transcription coupled repair. Our method of filtration differs from the
"rank-and-cut" method of [@PCAWG-MAIN] but represents a reasonable attempt to
account for the same biases. We also filtered genes on the basis of pathway
membership in biochemical pathways, excluding curated gene sets related to
diseases, syndromes, or classes of proteins with shared catalytic activity or
conserved domains ( _e.g._ transcription factors) which are potentially
problematic [@pathways-Atul-Butte]. We chose this approach to limit redundancy,
which is already high in the remaining pathways for some driver genes, and to
exclude biologically unrelated collections of genes. We do not account for
non-coding mutations, amplification events or structural variants, though we
hope to include them in subsequent studies (see below). Our pathway-centric
analysis also allows us to circumvent the problem that some cancer-promoting
mutations likely occur at such low frequencies as to be indistinguishable from
background. Previous attempts to capture these associations using Bayesian
networks have yielded important new insights into this problem and new genes
[@hornNetSigNetworkbasedDiscovery2018] that would not have been possible with
traditional analysis, supporting the case that pathway-centric approaches have
the potential to increase statistical power.

@@

An alternative hypothesis is that every tumor belongs to one of a large number
of syndromes which are unique to each tissue-of-origin, that share some
mechanisms and treatment strategies. Recent publication of TCGA consortium
papers present a view largely, and surprisingly, consistent with this hypothesis
[@hoadleyMultiplatformAnalysis2014; @hoadleyCellofOriginPatternsDominate2018].
Perhaps owing to the intractable complexity of genomics, proteomics, and
patient metadata in all its forms, the inescapable conclusion thus far is that
tissue-of-origin remains the most important driver of tumor characteristics at
every scale and by every measure. Our observations contrast with this view, and
instead support an interpretation of publicly available data in which all tumors
manifest one of a limited number of metastable phenotypes resulting from
disruptions in basic pathways as represented by our classes (**figures 1 & 2**).

@@

Our results illustrate how unique combinations of mutations in pan-cancer driver
genes with tissue-specific pathway disruptions result in common categories when
viewed at the level of the pathway knowledge-base. For example, we find that for
"snRNP assembly", one of the unique driver pathways of cluster 5, the top gene
in breast cancer is represented by TPR (mutated in 26% of samples). By contrast
the top gene for head and neck and lower grade gliomas is RANBP (20% and 13% of
samples, respectively). For some other clusters, _e.g._ 7, 8 & 9, the top genes
are known drivers such as TP53, PIK3CA and KRAS. None of these genes is 100%
mutated in any cluster however, and therefore it is the combinations with
secondary mutations, many of which are tissue-specific, that round out the
complement of pathways in all clusters. Some top genes are found in most of the
clusters (**supplemental figure S5**), in spite of the fact that they contribute to
many cluster-specific pathways. This can only be explained as a result of other
less common driver genes complementing the pathway disruption of the unique
combination of driver genes that are disrupted in each tumor, and we speculate
that many of these are relatively tissue-specific in terms of their sensitivity
to mutation. Consistent with our observations, @moonlight showed how tissue- and
cancer-specific driver genes and druggable targets of pathway moduli can be
discovered by machine learning on integrated multi-omics datasets. Marrying
these approaches in future work will greatly enhance our understanding of tumor
biology.

@@

By modeling CDC longevity data as a baseline risk function we showed that each
of our pathway disruption classes exhibit cancer-type specific effects on
expected survival (**figure 7**). Contrary to what one might expect at the
outset, we found that models assuming different survival based on class and
cancer type fit better than models in which they are independent factors.
However, considering that within each cancer type there are different clinical
standards of care, and even within classes of drugs the preferred treatment can
very between cancers, it makes sense that we observe tissue-specific class
effects. Contrast the situation with ovarian _vs._ breast cancer, which are both
hormonally driven cancers, for example. Ovarian cancer has but one main
treatment axis, platinum, whereas breast cancer patients have a variety of
treatment regimens based on molecular subtype and other factors. It is easy to
see how standard of care for one cancer and class could produce different
outcomes than the same class of a different cancer type. Unfortunately, given
the diversity of drug classes and treatments (including non-chemical treatments
such as surgery and palliative care), we lack sufficient power to explore these
variables in the TCGA data. It is our hope that future studies will help to
distinguish between treatment-specific effects on survival given different
pathway disruption classes.

### 2020-08-16 text/introduction.Rmd
modify spacing before and after all section & subsection headings in word style template; edit and reflow abstract and introduction
edit full document

Other studies aim at providing a comprehensive view of mutations, gene
expression and genomic signatures, with the express goal of understanding the
common themes of all cancer independent of tissue of origin. Understanding cancer
as a disease of the cell has long been a goal of
the field and is perhaps best characterized in essays by @weinberg2000; @hallmarks_II (see also
Discussion). The first of these studies in the enomic era identified 11 subtypes from 12 cancer types,
using integrative analysis with co-equal weighting of gene expression,
methylation, copy number and proteomics data
[@hoadleyMultiplatformAnalysis2014]. The principal finding was that
tissue-of-origin is the predominant driving factor, though a significant
proportion of tumors (~10%$) could be reclassified independent of
tissue-of-origin. In a second study involving 33 cancer types and a much greater
number of tumors [@shenIntegrativeClusteringMultiple2009], the authors identify 28 clusters that
could be further subdivided into organ specific groups, including
pan-gastrointestinal, pan-gynecological, pan-squamous, pan-gynecological/
squamous and pan-kidney [@hoadleyCellofOriginPatternsDominate2018].

More recently, attempts have been made to consider pan-cancer mutations and
other data outside the context of enrichment at the gene level. @PCAWG-PATH carried out near-exhaustive analysis of pathway enrichment using whole genome data. Horn _et al._
[@hornNetSigNetworkbasedDiscovery2018] introduce approaches to address the
sparseness of mutations, allowing the potential importance of a gene to be
influenced by the aggregate of mutations in its nearest network neighbors. This
method is powerful especially when applied to a small number (on the order of
hundreds) of tumors. One of the challenges in the field remains that beyond a
couple hundred common driver genes and mutations, there are likely thousands of
moderate effect genes that occur at such low frequency--given tissue diversity--and low sample numbers that they are impossible to detect using positive
selection theory. Some groups have attempted to address this challenge by
applying machine learning to cancer data sets to discover groups of functionally
related genes as they interact with larger pathways and networks
[@kimPancancerAnalysisSomatic2018, @mourikis-ml-pathways, @moonlight]. These
studies have proven very effective at highlighting some potentially fundamental
disease phenotypes at the pathway level that cross cancers with different
origins at the cellular and tissue level.

>>>>

Other studies aim at providing a comprehensive view of mutations, gene
expression and genomic signatures, with the express goal of understanding the
common themes of all cancer independent of tissue of origin. Understanding
cancer as a disease of the cell has long been a goal of the field and is perhaps
best characterized in essays by @weinberg2000; @hallmarks_II (see also
Discussion). The first of these studies in the genomic era identified 11 subtypes
from 12 cancer types, using integrative analysis with co-equal weighting of gene
expression, methylation, copy number and proteomics data
[@hoadleyMultiplatformAnalysis2014]. The principal finding was that
tissue-of-origin is the predominant driving factor, though a significant
proportion of tumors (~10%) could be reclassified independent of
tissue-of-origin. In a second study involving 33 cancer types and a much greater
number of tumors [@shenIntegrativeClusteringMultiple2009], the authors identify
28 clusters that could be further subdivided into organ specific groups,
including pan-gastrointestinal, pan-gynecological, pan-squamous,
pan-gynecological/ squamous and pan-kidney
[@hoadleyCellofOriginPatternsDominate2018].

More recently, attempts have been made to consider pan-cancer mutations and
other data as disruptions of normal pathway activity. @PCAWG-PATH
carried out near-exhaustive analysis of pathway enrichment using whole genome
data. Horn _et al._ [@hornNetSigNetworkbasedDiscovery2018] introduce approaches
to address the sparseness of mutations, allowing the potential importance of a
gene to be influenced by the aggregate of mutations in its nearest network
neighbors. This method is powerful especially when applied to a small number (on
the order of hundreds) of tumors. One of the challenges in the field remains
that beyond a couple hundred common driver genes and mutations, there are likely
thousands of moderate effect genes that occur at such low frequency--given
tissue diversity--and low sample numbers that they are impossible to detect
using positive selection theory. Some groups have attempted to address this
challenge by applying machine learning to cancer data sets to discover groups of
functionally related genes as they interact with larger pathways and networks
[@kimPancancerAnalysisSomatic2018; @mourikis-ml-pathways; @moonlight]. These
studies have proven very effective at highlighting some potentially fundamental
disease phenotypes at the pathway level that cross cancers with different
origins at the cellular and tissue level.

### 2020-08-16 text/manuscript.Rmd
edit full document

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, dpi=600, warning = FALSE)
suppressPackageStartupMessages(c(
  library(rstan),
  library(foreach),
  library(sqldf),
  library(tidyverse),
  library(ggthemes),
  library(data.table),
  library(biomaRt),
  library(dplyr),
  library(GenomicRanges),
  library(SummarizedExperiment),
  library(reactome.db),
  library(plyr),
  library(cluster),
  library(pheatmap),
  library(clusterProfiler),
  library(org.Hs.eg.db),
  library(dbscan),
  library(class),
  library(TCGAbiolinks),
  library(patchwork),
  library(ggplotify),
  library(ComplexHeatmap),
  library(knitr),
  library(kableExtra),
  library(UpSetR),
  library(cowplot),
  library(doRNG),
  library(uwot),
  library(janitor),
  library(FactoMineR),
  library(seriation),
  library(rtracklayer),
  library(poibin),
  library(flextable),
  library(plot3D),
  library(ggfortify)
))
```

@@

### A taxonomy of tumors based on disrupted molecular pathways
To study cancer pathways we obtained a set of 7,607 solid tumor samples from The
Cancer Genome Atlas (TCGA) through the [Genomic Data Commons
(GDC)](gdc.cancer.gov) portal. TCGA data were most appropriate for our
study given the relative completeness of the patient metadata, particularly for
survival and staging. We chose to analyze somatic mutations in exome
sequencing data because the affected target gene is known unambiguously.
Therefore, we selected all missense, nonsense, frameshift, stop-gain, untranslated
region, and splicing mutations. In order to minimize bias from well-studied diseases
and processes, we selected 377 Reactome pathways [reactome.org](https://reactome.org/) 
of interest corresponding to basic cellular processes and biochemical pathways,
excluding gene sets that correspond to miscellaneous categories ( _e.g._
“transcription factors”) or disease associations ( _e.g._ “mutated in colon
cancer”) and filtered our gene list on membership in these pathways (total
of 8940 genes; see methods for details).

To avoid bias toward larger
pathways ( _i.e._ pathways with more member genes), we counted pathways as disrupted if one or more member genes were mutated. We do not attempt to calculate enrichment for mutations within a pathway, for example. Data
binarized by pathway are likely to be noisy for several reasons. First, point
mutations can be deleterious (attenuating or hypomorphic) or activating
(neomorphic or hypermorphic) in genes, and these can in turn act functionally as either activators or repressors.
Within the scope of this study we accept this uncertainty and assume mutations
are generically disruptive to pathways regardless of whether they are
activating or deactivating or whether they affect a gene that we might classify
as one thing or another ( _e.g._ "Tumor suppressor" or "Oncogene"). Second, we
know that low-expressed genes and non-expressed genes are known to accumulate
mutations at a higher rate due to transcription coupled repair
[@kandothMutationalLandscapeSignificance2013; @kimPancancerAnalysisSomatic2018;
@Pervasive_lesion_segregation; @expression-aware-annotation]. To address this
issue, we identified genes with low expression in each type of cancer and
eliminated them for that cancer type only (see methods for details). 
Highly expressed genes could also have high mutation rates owing to
transcription induced mutagenesis [@parkGenomicEvidenceElevated2012]. We felt
that such mechanisms result in cell-type-specific biases that might be
biologically meaningful for predisposition to different classes of cancer
however, and therefore chose not to exclude these genes from our analysis. After
selecting our pathways and genes, we then compiled a matrix of the pathways,
assigning a Boolean value (a 1 or 0, indicating whether any pathway member gene
carries a mutation) to the pathway regardless of the number of genes carrying
mutations (**figure 1**).

In order to classify tumors using this dataset, we used multiple correspondence
analysis (MCA) [@JSSv025i01]. First, we determined the number of dimensions
containing useful information. (**figures 2A, B**). Then we performed a UMAP
analysis, both in order to summarize the MCA graphically, and as a preprocessing
step to boost the performance of density based clustering. The resulting map was
notable for its lobed structure, with several reproducible projections
regardless of random seed setting. A representative version of this
3D UMAP is shown in **figure 2C**, rotated to enhance the
visibility of the major features. Following this spatial mapping we attempted to
define groupings of similar tumors within the spatial map using HDBSCAN, which
performs hierarchical clustering and provides metrics of cluster stability and
probabilities of cluster membership for each node. However, HDBSCAN is sensitive
to several parameters; key for our analysis are the minimum number of tumor
samples in a cluster that capture the maximum number of tumor samples, measured
by probability of membership of ≥ 5% in at least one class. Thus, we
created a score metric as the fraction of classified tumors with max probability
< 5% in one class and chose a cluster size of 92 to minimize the score
function (**figure 2D**). HDBSCAN with these settings resulted in ten distinct
high-density clusters which we then projected onto the UMAP (**figure 2E**).
This classified 6,038 out of 7,607 tumors but still resulted in a
significant fraction of unclassified tumor samples. Since we ultimately wish to
be able to classify any tumor using this scheme, we performed k Nearest
Neighbors (kNN) analysis, which computes a similarity metric to every tumor in
the set and then lets the _k_ most similar tumors "vote" as to the identity of
the query tumor sample based on their classification labels. We set _k_ to be
the square root of the number of tumor samples (87). Using this method, we
assigned cluster membership to the remaining tumors (**figure 2F**).

@@

Many cancers have molecular or histological subtypes defined based on gene
expression or other -omics profiles or pathology lab results. These subtypes
often have different standards of care owing to different overall drug
sensitivity (or other factors). If the histological subtypes represent true
molecular phenotypes, one predicts that histological subtypes should segregate
with our pathway-based classes, therefore providing support for the classes as
proxies for molecular pathology sub-typing. To our surprise, we find a similar
result to the previous analysis of cancer types projected onto the UMAP of
pathway disruptions. To illustrate this, we projected annotations for each of the breast
cancer subtypes, composed of Triple-negative/Basal-like, Her2 positive,
normal-like, and luminal A and B subtypes onto the UMAP. These are among the
most heavily studied molecular subtypes in cancer, which each have different
prognoses and standards of care. We did not observe any exclusive segregation by
pathway for these subtype annotations (**Figure 3B** and **supplemental
file 1**). We also projected histological subtype data for the remaining cancers
(see **supplementary figure S2** for the full set of projections); we find that
the subtypes, though often biased towards one or more classes, are almost never
exclusive. We interpret these data in aggregate to mean
that our pathway disruption classes do not correspond to molecular subtypes
within the parent cancer type. Furthermore, this result raises the possibility
that, molecular and histo-pathological subtypes could represent different
precursor cell origins within each cancer type. This interpretation still leaves
open the possibility that different precursor cell types, being susceptible to
different teratogens and selection pressures ( _e.g._ insulin dependence) could
be biased toward one or more of the pathway disruption profiles.

@@

Finally, as a measure of tumor advancement, metastasis is the condition in which
certain phenotypic criteria are met. These phenotypes include (among others)
loss of differentiation, cell-cell contacts, epithelial to mesenchyme
transition, immune system evasion and tissue invasiveness [@hallmarks_II]. To determine whether
any of our classes correspond to an especially advanced stage of cancer across
tissue types, we projected the metastasis data onto our UMAP, and surprisingly we 
observed an even distribution of metastases across classes (**figure 3E**). This
final observation suggests that our pathway-disruption classification is
dependent on particular combinations of gene mutations that hit different
pathway profiles that can each give rise to advanced stages of disease and
metastasis, regardless and independent of overall mutational burden (see Discussion for further analysis).

@@

### A combination of common driver mutations and tissue-specific genes contribute to common class signatures

Prior efforts to extract common signatures from pan-cancer 
datasets met with difficulty in distinguishing tumor samples from their tissue-specific 
-omics data signatures. Given our pathway-disruption based classification, this raises 
the question, are tumor phenotypes driven entirely by common driver genes, or by “silent”
tissue-specific effectors ( _i.e._ too few samples to detect above statistical significance 
thresholds), or a combination of both? To answer this question, we compared top pathway genes for each cluster 
relative to the TCGA background to find differentially mutated genes. We ranked odds ratios
and selected the top ten enriched and depleted genes (pvalue < 0.01) for each cluster 
(**Figure 6**; odds ratios plot). Clusters 7 and 8, which shared multiple enrichment in 
signaling pathways, are largely driven by mutations in _PI3K_ and its homologs and _Ras_ 
genes, respectively (compare _PIK3CA_ and _KRAS_ panels of **Supplementary figure S5**). 
Interestingly, cluster 9, which also shared multiple enrichment in signaling pathways 
with clusters 7 and 8, is enriched for both _PIK3CA_ and _KRAS_. This result suggests 
that tumors in this cluster could be derivative of either cluster 7 or 8. Clusters 3 and 
5, defined by enrichment in metabolic pathways, had mutations in ribosomal proteins and 
nuclear pore complex, respectively. Cluster 4 had mutations in genes responsible for 
nucleosome structure. Cluster 2 had mutations in genes involved in protein degradation. 
We also observed that genes that were enriched for one cluster are depleted from others ( _e.g._ 
_TP53_ is enriched in cluster 6, but depleted in cluster 7; _PIK3CA_ is enriched in cluster
7, but depleted in clusters 3 and 8). Next, we investigated the proportion of samples per 
cancer type for the significant genes within a cluster (**Figure 6**; heatmap). Surprisingly, 
cluster-specific tumors were not predominated by one or more highly mutated genes across all 
cancers. Instead, when observing the mutation rate for these genes within samples that belong
to a cluster, the mutation rate is heterogeneous across tumors by tissue origin ( _e.g._ In 
cluster 4, CESC was enriched for _H2AFX_, OV was enriched for _HIST1H2BD_, and UCEC was enriched
for _HIST1H2AC_). Even among the top most mutationally enriched genes within clusters there is
no global pattern, indicating that our classes are not driven by individual genes, but rather 
networks as a whole. Taken together, our data identifies a framework of cancer type-specific 
mutations associated with specific clusters.

@@

The authors have nothing to declare. This work was supported

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, dpi=600, warning = FALSE)
suppressPackageStartupMessages(c(
  library(rstan),
  library(foreach),
  library(sqldf),
  library(tidyverse),
  library(ggthemes),
  library(data.table),
  library(biomaRt),
  library(dplyr),
  library(GenomicRanges),
  library(SummarizedExperiment),
  library(reactome.db),
  library(plyr),
  library(cluster),
  library(pheatmap),
  library(clusterProfiler),
  library(org.Hs.eg.db),
  library(dbscan),
  library(class),
  library(TCGAbiolinks),
  library(patchwork),
  library(ggplotify),
  library(ComplexHeatmap),
  library(knitr),
  library(kableExtra),
  library(UpSetR),
  library(cowplot),
  library(doRNG),
  library(uwot),
  library(janitor),
  library(FactoMineR),
  library(seriation),
  library(rtracklayer),
  library(poibin),
  library(flextable),
  library(plot3D)
))
```

@@

### A taxonomy of tumors based on disrupted molecular pathways

To study cancer pathways we obtained a set of 7,607 solid tumor samples from The
Cancer Genome Atlas (TCGA) through the [Genomic Data Commons
(GDC)](gdc.cancer.gov) portal. TCGA data were most appropriate for our study
given the relative completeness of the patient metadata, particularly for
survival and staging. We chose to analyze somatic mutations in exome sequencing
data because the affected target gene is known unambiguously. Therefore, we
selected all missense, nonsense, frameshift, stop-gain, untranslated region, and
splicing mutations. In order to minimize bias from well-studied diseases and
processes, we selected 377 [Reactome pathways](https://reactome.org/) of
interest corresponding to basic cellular processes and biochemical pathways,
excluding gene sets that correspond to miscellaneous categories ( _e.g._
“transcription factors”) or disease associations ( _e.g._ “mutated in colon
cancer”) and filtered our gene list on membership in these pathways (total of
8940 genes; see methods for details).

To avoid bias toward larger pathways ( _i.e._ pathways with more member genes),
we counted pathways as disrupted if one or more member genes were mutated. We do
not attempt to calculate enrichment for mutations within a pathway. Data
binarized by pathway are likely to be noisy for several reasons. First, point
mutations can be deleterious (attenuating or hypomorphic) or activating
(neomorphic or hypermorphic) in genes, and these can in turn act functionally as
either activators or repressors. Within the scope of this study we accept this
uncertainty and assume mutations are generically disruptive to normal pathway
activity regardless of whether they are activating or deactivating or whether
they affect a gene that we might classify as one thing or another ( _e.g._
"Tumor suppressor" or "Oncogene"). Second, we know that low-expressed genes and
non-expressed genes are known to accumulate mutations at a higher rate due to
transcription coupled repair [@kandothMutationalLandscapeSignificance2013;
@kimPancancerAnalysisSomatic2018;
@Pervasive_lesion_segregation; @expression-aware-annotation]. To address this
issue, we identified genes with low expression in each type of cancer and
eliminated them for that cancer type only (see methods for details). Highly
expressed genes could also have high mutation rates owing to transcription
induced mutagenesis [@parkGenomicEvidenceElevated2012]. We felt that such
mechanisms result in cell-type-specific biases that might be biologically
meaningful for predisposition to different classes of cancer however, and
therefore chose not to exclude these genes from our analysis. After selecting
our pathways and genes, we then compiled a matrix of the pathways, assigning a
Boolean value of 1 to each pathway with one or more genes mutated and 0 for all
others (**figure 1**).

In order to classify tumors using this dataset, we used multiple correspondence
analysis (MCA) [@JSSv025i01]. First, we determined the number of dimensions
containing useful information by selecting the eigenvalue with the most
explanatory power, using the average of 100 permutations of the data as baseline
(**figure 2A**). We then chose the maximum eigenvalue for which the p-value
remained ≤ 0.05 (see cutoff in **figure 2B**). Then we performed a UMAP
analysis, both in order to summarize the MCA graphically, and as a preprocessing
step to boost the performance of density based clustering. The resulting map was
notable for its lobed structure, with several reproducible projections
regardless of random seed setting. A representative version of this 3D UMAP is
shown in **figure 2C**, rotated to enhance the visibility of the major features.
Following this spatial mapping we attempted to define groupings of similar
tumors within the spatial map using HDBSCAN, which performs hierarchical
clustering and provides metrics of cluster stability and probabilities of
cluster membership for each node. However, HDBSCAN is sensitive to several
parameters; key for our analysis are the minimum number of tumor samples in a
cluster that capture the maximum number of tumor samples, measured by
probability of membership of ≥ 5% in at least one class. Thus, we created a
score metric as the fraction of classified tumors with max probability < 5% in
one class and chose a cluster size of 92 to minimize the score function
(**figure 2D**). HDBSCAN with these settings resulted in ten distinct
high-density clusters which we then projected onto the UMAP (**figure 2E**).
This classified 6,038 out of 7,607 tumors but still resulted in a significant
fraction of unclassified tumor samples. Since we ultimately wish to be able to
classify any tumor using this scheme, we performed k Nearest Neighbors (kNN)
analysis, which computes a similarity metric to every tumor in the set and then
lets the _k_ most similar tumors "vote" as to the identity of the query tumor
sample based on their classification labels. We set _k_ to be the square root of
the number of tumor samples (87). Using this method, we assigned cluster
membership to the remaining tumors (**figure 2F**).

@@

Many cancers have molecular or histological subtypes defined based on gene
expression or other -omics profiles or pathology lab results. These subtypes
often have different standards of care owing to different overall drug
sensitivity (or other factors). If the histological subtypes represent true
molecular phenotypes, one predicts that histological subtypes should segregate
with our pathway-based classes, therefore providing support for the classes as
proxies for molecular pathology sub-typing. To our surprise, we find a similar
result to the previous analysis of cancer types projected onto the UMAP of
pathway disruptions. To illustrate this, we projected annotations for each of
the breast cancer subtypes, composed of Triple-negative/Basal-like, Her2
positive, normal-like, and luminal A and B subtypes onto the UMAP. These are
among the most heavily studied molecular subtypes in cancer, which each have
different prognoses and standards of care. We did not observe any exclusive
segregation by pathway for these subtype annotations (**Figure 3B** and
**supplemental file 1**). We also projected histological subtype data for the
remaining cancers (see **supplementary figure S2** for the full set of
projections); we find that the subtypes, though often biased towards one or more
classes, are almost never exclusive. We interpret these data in aggregate to
mean that our pathway disruption classes do not correspond to previously
identified molecular subtypes within the parent cancer type. Furthermore, this
result raises the possibility that, molecular and histo-pathological subtypes
could represent different precursor cell origins within each cancer type. This
interpretation still leaves open the possibility that different precursor cell
types, being susceptible to different teratogens and selection pressures (
_e.g._ insulin dependence) could be biased toward one or more of the pathway
disruption profiles.

@@

Finally, as a measure of tumor advancement, metastasis is the condition in which
certain phenotypic criteria are met. These phenotypes include (among others)
loss of differentiation, cell-cell contacts, epithelial to mesenchyme
transition, immune system evasion and tissue invasiveness [@hallmarks_II]. To
determine whether any of our classes correspond to an especially advanced stage
of cancer across tissue types, we projected the metastasis data onto our UMAP,
and surprisingly we observed an even distribution of metastases across classes
(**figure 3E**). This final observation suggests that our pathway-disruption
classification is dependent on particular combinations of gene mutations
affecting different pathways that can each give rise to advanced stages of
disease and metastasis, regardless and independent of overall mutational burden
(see Discussion for further analysis).

@@

### A combination of driver mutations and tissue-specific genes contribute to common class signatures

Prior efforts to extract common signatures from pan-cancer datasets met with
difficulty in distinguishing tumor samples from their tissue-specific -omics
data signatures. Given our pathway-disruption based classification, this raises
the question, are tumor phenotypes driven entirely by common driver genes, or by
“silent” tissue-specific effectors ( _i.e._ too few samples to detect above
statistical significance thresholds), or a combination of both? To answer this
question, we compared top pathway genes for each cluster relative to the TCGA
background to find differentially mutated genes. We ranked odds ratios and
selected the top ten enriched and depleted genes (pvalue < 0.01) for each
cluster (**Figure 6**; odds ratios plot). Clusters 7 and 8, which shared
multiple enrichment in signaling pathways, are largely driven by mutations in
_PI3K_ and its orthologs and _Ras_ genes, respectively (compare _PIK3CA_ and
_KRAS_ panels of **Supplementary figure S5**). Interestingly, cluster 9, which
also shared multiple enrichment in signaling pathways with clusters 7 and 8, is
enriched for both _PIK3CA_ and _KRAS_. This result suggests that tumors in this
cluster could be derivative of either cluster 7 or 8. Clusters 3 and 5, defined
by enrichment in metabolic pathways, had mutations in ribosomal proteins and
nuclear pore complex, respectively. Cluster 4 had mutations in genes responsible
for nucleosome structure. Cluster 2 had mutations in proteasomal subunit genes
involved in protein degradation. We also observed that genes that were enriched
for one cluster are depleted from others ( _e.g. TP53_ is enriched in cluster
6, but depleted in cluster 7; _PIK3CA_ is enriched in cluster 7, but depleted in
clusters 3 and 8). Next, we investigated the proportion of samples per cancer
type for the significant genes within a cluster (**Figure 6**; heatmap).
Surprisingly, cluster-specific tumors were not predominated by one or more
highly mutated genes across all cancers. Instead, when observing the mutation
rate for these genes within samples that belong to a cluster, the mutation rate
is heterogeneous across tumors by tissue origin ( _e.g._ In cluster 4, CESC was
enriched for _H2AFX_, OV was enriched for _HIST1H2BD_, and UCEC was enriched for
_HIST1H2AC_). Even among the top most mutationally enriched genes within
clusters there is no global pattern, indicating that our classes are not driven
by individual genes, but rather networks as a whole. Taken together, our data
identifies a framework of cancer type-specific mutations associated with
specific clusters.

@@

The authors have nothing to declare. This work was supported by a grant from the Cedars-Sinai Precision Health Initiative.

### 2020-08-16 text/survival.Rmd
edit full document

To understand the model, we show in **Figure 7A** simulated survival time
curves for four related groups. We start with a cohort of women aged 30 and show
their expected survival (blue). By setting their cancer rate multiplier to $k=2$
we can simulate the effect of a moderately deadly cancer (red). Compare this to
a randomly selected group of women ranging in age from 30 to 70 years old (dark
blue). Immediately the survival curve changes due to the mixture of ages,
without any malignancy. Adding a $k=2$ malignancy further reduces expected
lifespan (dark red). We can see that for different tissues, simply due to the
change in distribution of age-at-diagnosis, we should expect equally deadly
cancers to have *different* survival curves. The model takes this into account
and the aggressiveness of the cancer can be estimated without the confounding
aspect of age.

>>>>

To understand the model, we show in **Figure 7A** simulated survival time curves
for four related groups. We start with a cohort of women aged 30 and show their
expected survival (light blue). By setting their cancer rate multiplier to $k=2$
we can simulate the effect of a moderately deadly cancer (yellow). Compare this
to a randomly selected group of women ranging in age from 30 to 70 years old
(dark blue). Immediately the survival curve changes due to the mixture of ages,
without any malignancy. Adding a $k=2$ malignancy further reduces expected
lifespan (red). We can see that for different tissues, simply due to the change
in distribution of age-at-diagnosis, we should expect equally deadly cancers to
have *different* survival curves. The model takes this into account and the
aggressiveness of the cancer can be estimated without the confounding aspect of
age.
